Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-07-31 manuscripts/reply/reply-acknowledgments.tex
More editing of reply ms.
Changing discussions of "estimation error" to "numerical
approximation error" as per Vladimir's comments. Adding Vlad
to the acknowledgments as well.

We thank Melissa Callahan, Jake Esselstyn, Cameron Siler, Mark Holder, Rafe
Brown, Emily McTavish, Daniel Money, Jordan Koch, and Adam Leach\'{e} for
insightful comments that greatly improved this work.
We thank Michael Hickerson and co-authors for generously providing their data.
J.\ Oaks and C.\ Linkem thank the National Science Foundation for supporting
this work (DEB 1011423, DBI 1308885 and BIO-1202754).
J.\ Oaks was also supported by the University of Kansas (KU) Office of Graduate
Studies, Society of Systematic Biologists, Sigma Xi Scientific Research
Society, KU Department of Ecology and Evolutionary Biology, and the KU
Biodiversity Institute.
We also thank Mark Holder, the KU Information and Telecommunication Technology
Center, KU Computing Center, and the iPlant Collaborative for the computational
support necessary to conduct the analyses presented herein.

>>>>

We thank Melissa Callahan, Jake Esselstyn, Cameron Siler, Mark Holder, Rafe
Brown, Emily McTavish, Daniel Money, Jordan Koch, Adam Leach\'{e}, Vladimir
Minin, Luke Harmon, and three anonymous reviewers for insightful comments that
greatly improved this work.
We thank Michael Hickerson and co-authors for generously providing their data.
J.\ Oaks and C.\ Linkem thank the National Science Foundation for supporting
this work (DEB 1011423, DBI 1308885 and BIO-1202754).
J.\ Oaks was also supported by the University of Kansas (KU) Office of Graduate
Studies, Society of Systematic Biologists, Sigma Xi Scientific Research
Society, KU Department of Ecology and Evolutionary Biology, and the KU
Biodiversity Institute.
We also thank Mark Holder, the KU Information and Telecommunication Technology
Center, KU Computing Center, and the iPlant Collaborative for the computational
support necessary to conduct the analyses presented herein.

### 2014-07-31 manuscripts/reply/reply-main.tex
More editing of reply ms.
Changing discussions of "estimation error" to "numerical
approximation error" as per Vladimir's comments. Adding Vlad
to the acknowledgments as well.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method was very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       spurious support for models with few divergence events shared across
       taxa. Thus, the primary inference enabled by the approach is very
       sensitive to the priors on divergence times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{itemize}
    \item \citet{Hickerson2013} suggest that under broad uniform priors, the
        rejection algorithm implemented in \msb is unable to adequately sample
        the space of the models within reasonable computational time, which
        leads to support for models with fewer divergence-time parameters
        because they are better sampled.
    \item \citet{Oaks2012} suggest the cause is more fundamental; the uniform
        priors on divergence times simply lead to very small marginal
        likelihoods of models with many divergence-time parameters. These
        models are ``averaged'' over much greater parameter space with large
        prior weight and small likelihood \citep{Jeffreys1939,Lindley1957}.
\end{itemize}
In the former case, the problem is estimation error; the true (correct)
posterior supports models with many divergence-time parameters given data
generated under temporally random divergences, but we are unable to accurately
estimate the posterior.
In the latter case, estimation error is not the issue, because the true
posterior supports simultaneous divergences given data from randomly diverged
taxa.
I.e., this is the correct (albeit biogeographically misleading) answer given
the prior assumption that divergence times are broadly and uniformly
distributed.
Such posterior support for simultaneous divergence, even if correct from the
perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

While these phenomena are not mutually exclusive, it is important to
distinguish between them in order to determine how best to improve our ability
to estimate shared divergence histories.
If the cause is estimation error, then the model is sound and we need to
improve our numerical estimation procedures.
For example, Markov chain or sequential Monte Carlo algorithms might improve
the efficiency of sampling from the posterior.
Rather than alter the sampling algorithm, \citet{Hickerson2013} tried a
model-averaging approach that uses narrow, empirically informed uniform priors
in the hope that with less parameter space to sample, the rejection algorithm
will produce better estimates of the posterior.
Here, we discuss theoretical considerations for using empirically informed
priors for Bayesian model choice and evaluate the approach of
\citet{Hickerson2013} as a potential solution to the biases of \msb revealed by
\citet{Oaks2012}.
In their analyses, \citet{Hickerson2013} made an error by mixing different
units of time, which invalidates the results presented in their response (see
Supporting Information for details).
We correct this error, but still find their approach will often support
(1) clustered divergence models when divergences are random, and
(2) models that exclude from consideration the true values of the
parameters.
The behavior of their approach also suggests that estimation error is not the
primary cause of the issue.

If support for models with few divergence events is caused by their larger
marginal likelihoods, we need to correct the model; improving the estimation
machinery will only increase our ability to estimate misleading posteriors.
Accordingly, \citet{Oaks2014dpp} has introduced a method that uses more
flexible probability distributions to accommodate prior uncertainty in
divergence times without overly inhibiting the marginal likelihoods of models
with more divergence-time parameters.
This greatly increases the method's robustness and power to detect temporal
variation in divergences \citep{Oaks2014dpp}.
This is not surprising given the rich statistical literature showing that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection
\citep[e.g.,][]{Jeffreys1939,Lindley1957}.

We also explore the distinct predictions made by the alternative causes of the
bias proposed by \citet{Oaks2012} and \citet{Hickerson2013}.
We show the primary cause of the method's tendency to support models of shared
divergences is the large marginal likelihoods of these models under the prior
assumption of uniformly distributed divergence times.
We also show that sampling error is not a major contributer to the problem.

@@

However, \citet{Hickerson2013} conclude that the bias is caused by estimation
error due to insufficient sampling.
They argue the widest of the three priors used by \citet{Oaks2012} would
infrequently produce samples with many independent population divergence times
as recent as the estimated gene divergence times presented in \citet{Oaks2012}.
However, this argument assumes the gene divergence times presented in
\citet{Oaks2012} were estimated without error.
These estimates were intended to provide only a rough comparison of the
gene divergence times across the 22 taxa.
These analyses assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations
per generation for all taxa, and are, of course, subject to estimation error.
Furthermore, the branch-length units of the gene trees are in millions of
years, whereas the divergence-time prior of \msb is in generations, thus
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
The argument of \citet{Hickerson2013} that divergence-time estimates of
\citet{Oaks2012} should set an upper bound on the prior for divergence times
seems questionable, especially given our findings presented above regarding the
behavior of \msb when empirically informed uniform priors are employed.

@@

To disentangle whether model likelihoods or insufficient prior sampling is to
blame for the method's erroneous support for simple models, we must look at the
different predictions made by these two phenomena.
For example, insufficient prior sampling should create large variance among
posterior estimates and cause analyses to be highly sensitive to the number of
samples drawn from the prior.
Furthermore, if sampling error is \emph{biasing} estimates toward models with
less parameter space, as suggested by \citet{Hickerson2013}, we expect to see
support for these models decrease as sampling increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~S\ref{figSamplingError}) we do not see the trend predicted by
\citet{Hickerson2013}.
While sampling error is always a reality, it does not appear to be playing a
large role in the biases revealed by the results of \citet{Oaks2012} or
presented above.

A straightforward prediction if marginal likelihoods are causing the preference
for simple models is that the bias should disappear as the model generating the
data converges to the prior.
\citet{Oaks2012} tested this prediction by performing 100,000 simulations to
assess the model-choice behavior of \msb when the prior model is correct.
The results confirm the prediction:
\msb estimates the probability of the one-divergence model quite well when the
prior is correct (see Figure 4 of \citet{Oaks2012}).
We confirmed this same behavior for the model-averaging approach used by
\citet{Hickerson2013} (see SI text and Figure~S\ref{figValidationMCBehavior}).
These results are not clearly predicted if insufficient sampling was causing
estimation bias.
Even when the prior is correct, due to the discrete uniform prior on the number
of divergence events (\numt{}) implemented in \msb, models with larger numbers
of divergence-time parameters (and thus greater parameter space) will still be
far less densely sampled than those with fewer divergence events
\citep{Oaks2012}.
Thus, the results of the simulations of \citet{Oaks2012} are more consistent
with the fundamental sensitivity of marginal likelihoods to priors.

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and fewer \divt{} parameters (Figure~\ref{figPower4}).
In all of these analyses, each of the prior models have the same number of
samples.
Thus, while sampling error will always be present in any numerical Bayesian
approximation method, insufficient sampling is an unlikely explanation for the
erroneous support for models with less parameter space.
While analyses that sample each model proportional to their parameter space
could be explored, it is clear that the marginal likelihoods under broad
uniform priors on divergence times will be greater for models with fewer
divergence-time dimensions.

@@

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that may exclude the true values
of the parameters.
Our results show that it is difficult to choose an uniformly distributed
prior on divergence times that is broad enough to confidently contain the true
values of parameters while being narrow enough to avoid erroneous support for
models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection; these prior assumptions strongly
influence the posterior probabilities we seek to estimate.

>>>>

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       spurious support for models with few divergence events shared across
       taxa. Thus, the primary inference enabled by the approach is very
       sensitive to the priors on divergence times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{itemize}
    \item \citet{Hickerson2013} suggest that under broad uniform priors, the
        rejection algorithm implemented in \msb is unable to adequately sample
        the space of the models within reasonable computational time, which
        leads to bias toward models with fewer divergence-time parameters
        because they are better sampled.
    \item \citet{Oaks2012} suggest the cause is more fundamental; the uniform
        priors on divergence times simply lead to very small marginal
        likelihoods of models with many divergence-time parameters. These
        models are ``averaged'' over much greater parameter space with large
        prior weight and small likelihood \citep{Jeffreys1939,Lindley1957}.
\end{itemize}
In the former case, the problem is numerical approximation error due to
insufficient computation;
the true (correct) posterior supports models with many divergence-time
parameters given data generated under temporally random divergences, but we are
unable to accurately approximate the posterior.
In the latter case, numerical error is not the issue, because the true
posterior supports simultaneous divergences given data from randomly diverged
taxa.
I.e., this is the correct (albeit biogeographically misleading) answer given
the prior assumption that divergence times are broadly and uniformly
distributed.
Such posterior support for simultaneous divergence, even if correct from the
perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

While these phenomena are not mutually exclusive, it is important to
distinguish between them in order to determine how best to improve our ability
to estimate shared divergence histories.
If the cause is numerical approximation error, then the model is sound and we
need to increase our computational effort or improve our Monte Carlo
integration procedures.
For example, Markov chain or sequential Monte Carlo algorithms might sample the
posterior more efficiently than the simple Monte Carlo rejection sampler
currently implemented in \msb.
Rather than alter the sampling algorithm, \citet{Hickerson2013} tried a
model-averaging approach that uses narrow, empirically informed uniform priors
in the hope that with less parameter space to sample, the rejection algorithm
will produce better estimates of the posterior.
Here, we discuss theoretical considerations for using empirically informed
priors for Bayesian model choice and evaluate the approach of
\citet{Hickerson2013} as a potential solution to the biases of \msb revealed by
\citet{Oaks2012}.
In their analyses, \citet{Hickerson2013} made an error by mixing different
units of time, which invalidates the results presented in their response (see
Supporting Information for details).
We correct this error, but still find their approach will often support
(1) clustered divergence models when divergences are random, and
(2) models that exclude from consideration the true values of the
parameters.
The behavior of their approach also suggests that insufficient computation is
not the primary cause of the issue.

If support for models with few divergence events is caused by their larger
marginal likelihoods, we need to correct the model; improving the numerical
approximation machinery will only increase our ability to estimate the true
posterior that leads us to the wrong interpretation about evolutionary history.
Accordingly, \citet{Oaks2014dpp} has introduced a method that uses more
flexible probability distributions to accommodate prior uncertainty in
divergence times without overly inhibiting the marginal likelihoods of models
with more divergence-time parameters.
This greatly increases the method's robustness and power to detect temporal
variation in divergences \citep{Oaks2014dpp}.
This is not surprising given the rich statistical literature showing that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection
\citep[e.g.,][]{Jeffreys1939,Lindley1957}.

We also explore the distinct predictions made by the alternative causes of the
bias proposed by \citet{Oaks2012} and \citet{Hickerson2013}.
We show the primary cause of the method's tendency to support models of shared
divergences is the large marginal likelihoods of these models under the prior
assumption of uniformly distributed divergence times.
We also show that numerical approximation error is not a major contributer to
the problem.

@@

However, \citet{Hickerson2013} conclude that the bias is caused by
numerical approximation error due to insufficient computation.
They argue the widest of the three priors used by \citet{Oaks2012} would
infrequently produce random samples with many independent population divergence
times as recent as the estimated gene divergence times presented in
\citet{Oaks2012}.
However, this argument assumes the gene divergence times presented in
\citet{Oaks2012} were estimated without error.
These estimates were intended to provide only a rough comparison of the
gene divergence times across the 22 taxa.
These analyses assumed an arbitrary strict per-site rate of $2\e{-8}$ mutations
per generation for all taxa, and are, of course, subject to estimation error.
Furthermore, the branch-length units of the gene trees are in millions of
years, whereas the divergence-time prior of \msb is in generations, thus
\citet{Hickerson2013} make the implicit assumption that all 22 Philippine taxa
have a generation time of one year.
The argument of \citet{Hickerson2013} that divergence-time estimates of
\citet{Oaks2012} should set an upper bound on the prior for divergence times
seems questionable, especially given our findings presented above regarding the
behavior of \msb when empirically informed uniform priors are employed.

@@

To disentangle whether model likelihoods or numerical approximation error is to
blame for the method's erroneous support for simple models, we must look at the
different predictions made by these two phenomena.
For example, numerical error due to insufficient prior sampling should create
large variance among posterior estimates and cause analyses to be highly
sensitive to the number of samples drawn from the prior.
Furthermore, if insufficient prior sampling is \emph{biasing} estimates toward
models with less parameter space, as suggested by \citet{Hickerson2013}, we
expect to see support for these models decrease as sampling from the prior
increases.
\citet{Oaks2012} did not see such sensitivity when they compared prior sample
sizes of $2\e6$, $5\e6$, and $10^7$.

To explore this prediction further, we repeat the analysis of the Philippines
dataset under the intermediate prior used by \citet{Oaks2012} ($\divt{} \sim
U(0, 10)$, $\meanDescendantTheta{} \sim (0.0005, 0.04)$, $\ancestralTheta{}
\sim (0.0005, 0.02)$), using a very large prior sample size of $10^8$.
When we look at the trace of the estimates of the dispersion index of
divergence times (\vmratio{}) as the prior samples accumulate
(Figure~S\ref{figSamplingError}) we do not see the trend predicted by
\citet{Hickerson2013}.
While approximation error is always present for any numerical analysis, it does
not appear to be playing a large role in the biases revealed by the results of
\citet{Oaks2012} or presented above.

A straightforward prediction if marginal likelihoods are causing the preference
for simple models is that the bias should disappear as the model generating the
data converges to the prior.
\citet{Oaks2012} tested this prediction by performing 100,000 simulations to
assess the model-choice behavior of \msb when the prior model is correct.
The results confirm the prediction:
\msb estimates the probability of the one-divergence model quite well when the
prior is correct (see Figure 4 of \citet{Oaks2012}).
We confirmed this same behavior for the model-averaging approach used by
\citet{Hickerson2013} (see SI text and Figure~S\ref{figValidationMCBehavior}).
These results are not clearly predicted if insufficient computation was causing
numerical error.
Even when the prior is correct, due to the discrete uniform prior on the number
of divergence events (\numt{}) implemented in \msb, models with larger numbers
of divergence-time parameters (and thus greater parameter space) will still be
far less densely sampled than those with fewer divergence events
\citep{Oaks2012}.
Thus, the results of the simulations of \citet{Oaks2012} are more consistent
with the fundamental sensitivity of marginal likelihoods to priors.

This is further demonstrated by the results presented herein that show the
model-averaging approach of \citet{Hickerson2013} prefers models with narrower
\divt{} priors (Table~\ref{tabModelChoiceEmpirical} and
Figs.~\labelcref{figExclusionSimTau,figExclusionSimProb,figPowerExclusion4})
and fewer \divt{} parameters (Figure~\ref{figPower4}).
In all of these analyses, the same number of random samples were drawn from
each of the prior models.
Thus, while approximation error will always be present in any numerical
approximation method, insufficient prior sampling is an unlikely explanation
for the erroneous support for models with less parameter space.
While analyses that sample each model proportional to their parameter space
could be explored, it is clear that the marginal likelihoods under broad
uniform priors on divergence times will be greater for models with fewer
divergence-time dimensions.

@@

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that may exclude the true values
of the parameters.
Our results show that it is difficult to choose an uniformly distributed
prior on divergence times that is broad enough to confidently contain the true
values of parameters while being narrow enough to avoid erroneous support for
models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection; these prior assumptions about
parameters strongly influence the posterior probabilities we seek to estimate.
